D3064R0: How to Avoid OOTA Without Really Trying

### 2024-02-29 ootavacuous.tex
Edits for Appendices B and C
Here is a bunch of small changes to Appendices B and C.  The largest set
of changes involves altering the line breaks in the GCC command-line
example to prevent overfull lines.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

As noted earlier, the exact definition of a computer language is subject
to some debate, with standards, implementations, and users all having
some degree of
influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2},
and each of which is subject to change over time.
It is natural to dismiss user influence when compared to the text of
standards or the code in implementations, but both are subject to
change and do change over time.
The implementation is especially subject to change, for example,
consider the following command line used to build the Linux kernel's
\co{kernel/rcu/tree.c} C-language source file:

~\\
{
	\scriptsize
	\texttt{
	gcc -Wp,-MMD,kernel/rcu/.tree.o.d -nostdinc
	-I./arch/x86/include \\
	-I./arch/x86/include/generated  -I./include
	-I./arch/x86/include/uapi \\
	-I./arch/x86/include/generated/uapi
	-I./include/uapi -I./include/generated/uapi \\
	-include ./include/linux/compiler-version.h
	-include ./include/linux/kconfig.h \\
	-include ./include/linux/compiler\_types.h -D\_\_KERNEL\_\_
	-fmacro-prefix-map=./= \\
	-Werror -std=gnu11 -fshort-wchar
	-funsigned-char -fno-common -fno-PIE \\
	-fno-strict-aliasing -mno-sse
	-mno-mmx -mno-sse2 -mno-3dnow -mno-avx \\
	-fcf-protection=branch
	-fno-jump-tables -m64 -falign-jumps=1 -falign-loops=1 \\
	-mno-80387 -mno-fp-ret-in-387 -mpreferred-stack-boundary=3
	-mskip-rax-setup \\
	-mtune=generic -mno-red-zone -mcmodel=kernel
	-Wno-sign-compare \\
	-fno-asynchronous-unwind-tables
	-mindirect-branch=thunk-extern \\
	-mindirect-branch-register
	-mindirect-branch-cs-prefix -mfunction-return=thunk-extern
	-fno-jump-tables -fpatchable-function-entry=16,16
	-fno-delete-null-pointer-checks -O2 \\
	-fno-allow-store-data-races
	-fstack-protector-strong -fomit-frame-pointer \\
	-fno-stack-clash-protection -falign-functions=16
	-fno-strict-overflow -fno-stack-check \\
	-fconserve-stack
	-Wall -Wundef -Werror=implicit-function-declaration \\
	-Werror=implicit-int -Werror=return-type -Werror=strict-prototypes \\
	-Wno-format-security -Wno-trigraphs -Wno-frame-address
	-Wno-address-of-packed-member \\
	-Wframe-larger-than=2048 -Wno-main
	-Wno-unused-but-set-variable \\
	-Wno-unused-const-variable -Wvla
	-Wno-pointer-sign -Wcast-function-type \\
	-Wno-array-bounds
	-Wno-alloc-size-larger-than -Wimplicit-fallthrough=5 \\
	-Werror=date-time -Werror=incompatible-pointer-types
	-Werror=designated-init \\
	-Wenum-conversion
	-Wno-unused-but-set-variable -Wno-unused-const-variable \\
	-Wno-restrict -Wno-packed-not-aligned -Wno-format-overflow
	-Wno-format-truncation \\
	-Wno-stringop-overflow
	-Wno-stringop-truncation -Wno-missing-field-initializers \\
	-Wno-type-limits -Wno-shift-negative-value
	-Wno-maybe-uninitialized \\
	-Wno-sign-compare
	-DKBUILD\_MODFILE='"kernel/rcu/tree"'
	-DKBUILD\_BASENAME='"tree"' \\
	-DKBUILD\_MODNAME='"tree"' -D\_\_KBUILD\_MODNAME=kmod\_tree \\
	-c -o kernel/rcu/tree.o kernel/rcu/tree.c
	}
}

@@

The \co{-funsigned-char} causes type \co{char} to be unsigned, which
overrides per-architecture defaults, some of which treat \co{char}
as signed and others as \co{unsigned}.
This choice prevents a class of bugs, and also allows the kernel to make
reliable use of the uppermost bit of variables of type \co{char}.
It also affects the definition of ``semantic dependency'' by changing
the arithmetic proporties of this type.
In theory, the standard could specify the signedness of \co{char}, but the
variety of existing practice in the 1980s prevented this.
Plus machine instructions were much more expensive back then, so the
signedness of \co{char} was much more of a performance concern than
it is toda than
it is todayy.

The \co{-mno-sse} prevents GCC from making use of the SSE hardware.
This is done for performance reasons, as it avoids the overhead of
saving and restoring the state of this hardware when switching between
user and kernel contexts.
Similarly, the \co{-mcmodel=kernel} causes the kernel binary to be
placed in the uppermost 2GB of the address space, again reducing
the overhead of switching between user and kernel contexts.
These are cases where the standard does not specify anything, nor
should it.

The \co{-fpatchable-function-entry=16,16} causes GCC to emit 16
\co{nop} instructions at the beginning of each function, with
the function's entry point being just after this string of \co{nop}
instructions.
The resulting buffers are used by the Linux-kernel tracing infrastructure
which is in turn used for debugging, performance measurement, and
monitoring.
This is again clearly outside of the scope of the standard.

The \co{-fstack-protector-strong} causes GCC to emit code that provides
some protection against some clases of attacks based on buffer overflows.
One could rightly argue code should simply avoid ever overflowing buffers,
but things like memory allocators and userspace memory accesses must
use code that can be difficult to distinguish from buffer overflows.
It is not clear that the ever-increasing variety of attacks should
affect the standard.

The \co{-fno-strict-overflow} causes GCC to act as if signed integer
overflow is defined behavior.
Note well that this command-line option also affects the definition of
``semantic dependency''.
This might be a controversial choice, and another option would be to add
a new set of signed integer types to the standard for which overflow is
defined as wrapping, similar to the situation with unsigned integers.

The \co{-Werror=strict-prototypes} causes GCC to warn if old-style
non-ANSI function prototypes are used.
This helps avoid certain classes of bugs.
Warnings are by design outside of the scope of the standard.

In other words, user preference can exert a nontrivial influence over
language semantics, and in particular can affect some aspects of the
definition of ``semantic dependency''.

@@

One entirely reasonable reaction is that, given the issues raised in
Sections~\ref{sec:Properties of Semantic Dependencies}
and~\ref{sec:The Fundamental Property of Semantic Dependencies},
load/store reordering is the least of the problems faced by tooling.
Nevertheless, this appendix expands on
Section~\ref{sec:Semantic Dependencies and Tooling}
by looking into the possibility of:

\begin{enumerate}
\item	Tooling focusing on only part of the language,
\item	Changing the language to eliminate some aspects that are
	troublesome for tooling (and potentially accepting performance
	and energy-efficiency shortfalls), and
\item	Changing the language to better delineate those portions that
	tooling can easily accommodate.
\end{enumerate}

@@

Both the ARM and PowerPC memory models permit reordering prior loads
against subsequent stores, but actual tests on recent hardware fail to
produce any evidence of any such reordering.
Nevertheless, thus far neither ARM nor PowerPC hardware architects
have been willing to strengthen their memory models so as to forbid
such reordering.
% @@@ Cite Grisenthwaite

@@

However, if that load suffers a cache miss, address translation will have
completed long before the load returns its value, meaning that a later
store might well commit before the load completes.
In fact, that later store might commit before the store that supplies the
value returned by that load!
Which explains why some hardware systems really do reorder prior loads
and later stores.

@@

However, some kernels and applications have ways of handling even
uncorrectable errors.
Operating-system kernels encountering an ECC error might note that the
affected data was being used by only one user process, and might react by
killing that user process, taking care to account for memory shared with
other processes.
Some user applications might be note that the affected data affected only
a particular computation, and might react by restarting that computation.
In both cases, it is only reasonable to react to an uncorrectable ECC
error in a load from a read-only mapped file by re-reading that data
from that file, then restarting that load instruction.
In all cases, the kernel and the user application would need to continue
execution, and would thus require an exact exception.

@@

If you would like an authoratative answer to this question, you should of
course ask your friendly local hardware architect.
In the meantime, here is some semi-informed speculation on this topic:

\begin{enumerate}
\item	Some hardware might choose to forego ECC, for example, in order to
	reduce cost for low-end systems or to improve energy efficiency
	for battery-powered systems.
\item	ECC error correction might still be done in hardware, avoiding
	the need for precise software-visible exceptions.
\item	Some systems might prefer to immediately shut down in response to
	an uncorrectable ECC error, perhaps due to safety considerations.
	This would entirely avoid the need for software-visible
	ECC-related exceptions.\footnote{
		One motivation for immediate shutdown is that if there
		are uncorrectable errors, there might soon be errors
		that translate one valid bit pattern to another valid
		bit pattern, which could result in a lack of safety.}
\item	Someone might come up with a clever way of correcting ECC errors
	in firmware, avoiding the need for precise software-visible
	exceptions.
\item	As late as early 2024, some GPGPUs have been observed reordering
	earlier loads against later stores.
	Other hardware architects might therefore feel the need to keep
	this option open for their own systems.
\end{enumerate}

@@

However, this would mean that programs compiled in this way would not
be guaranteed to run correctly on weakly ordered systems that lack
ECC (or that lack precise exceptions).
In addition, this change is more strict than necessary, forbidding
optimizations that tooling could in fact analyze.
Therefore, the next section looks at standardizaing a less severe
restriction.

>>>>

As noted earlier, the exact definition of a computer language is subject
to some debate, with standards, implementations, and users all having
some degree of
influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2}.
It is natural to dismiss user influence when compared to the text of
standards or the code in implementations, but both are subject to
change and do change over time.
An especially easy way for users to influence the implementation is by
means of command-line flags and switch settings.
As an example,
consider the following command line used to compile the Linux kernel's
\co{kernel/rcu/tree.c} C-language source file:

\begin{quote}
{
	\scriptsize
	\texttt{\noindent
	gcc -Wp,-MMD,kernel/rcu/.tree.o.d -nostdinc
	-I./arch/x86/include \\
	-I./arch/x86/include/generated  -I./include
	-I./arch/x86/include/uapi \\
	-I./arch/x86/include/generated/uapi
	-I./include/uapi \\
	-I./include/generated/uapi \\
	-include ./include/linux/compiler-version.h \\
	-include ./include/linux/kconfig.h \\
	-include ./include/linux/compiler\_types.h \\
	-D\_\_KERNEL\_\_
	-fmacro-prefix-map=./=
	-Werror -std=gnu11 -fshort-wchar \\
	-funsigned-char -fno-common -fno-PIE
	-fno-strict-aliasing \\
	-mno-sse
	-mno-mmx -mno-sse2 -mno-3dnow -mno-avx \\
	-fcf-protection=branch
	-fno-jump-tables -m64 -falign-jumps=1 \\
	-falign-loops=1 -mno-80387 -mno-fp-ret-in-387 \\
	-mpreferred-stack-boundary=3 -mskip-rax-setup \\
	-mtune=generic -mno-red-zone -mcmodel=kernel
	-Wno-sign-compare \\
	-fno-asynchronous-unwind-tables
	-mindirect-branch=thunk-extern \\
	-mindirect-branch-register
	-mindirect-branch-cs-prefix \\
	-mfunction-return=thunk-extern -fno-jump-tables \\
	-fpatchable-function-entry=16,16
	-fno-delete-null-pointer-checks \\
	-O2 -fno-allow-store-data-races
	-fstack-protector-strong \\
	-fomit-frame-pointer
	-fno-stack-clash-protection -falign-functions=16 \\
	-fno-strict-overflow -fno-stack-check -fconserve-stack
	-Wall -Wundef \\
	-Werror=implicit-function-declaration
	-Werror=implicit-int \\
	-Werror=return-type
	-Werror=strict-prototypes -Wno-format-security \\
	-Wno-trigraphs -Wno-frame-address \\
	-Wno-address-of-packed-member
	-Wframe-larger-than=2048 -Wno-main \\
	-Wno-unused-but-set-variable -Wno-unused-const-variable -Wvla \\
	-Wno-pointer-sign -Wcast-function-type -Wno-array-bounds \\
	-Wno-alloc-size-larger-than -Wimplicit-fallthrough=5 \\
	-Werror=date-time -Werror=incompatible-pointer-types \\
	-Werror=designated-init -Wenum-conversion
	-Wno-unused-but-set-variable \\
	-Wno-unused-const-variable -Wno-restrict -Wno-packed-not-aligned \\
	-Wno-format-overflow 
	-Wno-format-truncation -Wno-stringop-overflow \\
	-Wno-stringop-truncation
	-Wno-missing-field-initializers \\
	-Wno-type-limits
	-Wno-shift-negative-value -Wno-maybe-uninitialized \\
	-Wno-sign-compare 
	-DKBUILD\_MODFILE='"kernel/rcu/tree"' \\
	-DKBUILD\_BASENAME='"tree"' -DKBUILD\_MODNAME='"tree"' \\
	-D\_\_KBUILD\_MODNAME=kmod\_tree \\
	-c -o kernel/rcu/tree.o kernel/rcu/tree.c
	}
}
\end{quote}

@@

The \co{-funsigned-char} causes the \co{char} type to be unsigned, which
overrides per-architecture defaults, some of which treat \co{char}
as signed and others as \co{unsigned}.
This choice prevents a class of bugs, and also allows the kernel to make
reliable use of the uppermost bit of variables of type \co{char}.
It also affects the definition of ``semantic dependency'' by changing
the arithmetic proporties of this type.
In theory, the standard could have specified the signedness of \co{char}
but the variety of existing practice in the 1980s prevented this.
Plus machine instructions were much more expensive back then, so the
signedness of \co{char} was much more of a performance concern than
it is today.

The \co{-mno-sse} prevents GCC from making use of the processor's SSE hardware.
This is done for performance reasons, as it avoids the overhead of
saving and restoring the state of this hardware when switching between
user and kernel contexts.
Similarly, the \co{-mcmodel=kernel} causes the kernel binary to be
placed in the uppermost 2GB of the address space, again reducing
the overhead of switching between user and kernel contexts.
These are cases where the standard does not specify anything, nor
should it.

The \co{-fpatchable-function-entry=16,16} causes GCC to emit 16
\co{nop} instructions at the beginning of each function, with
the function's entry point being just after this string of \co{nop}
instructions.
The resulting buffers are used by the Linux-kernel tracing infrastructure
which is in turn used for debugging, performance measurement, and
monitoring.
This is again clearly outside the scope of the standard.

The \co{-fstack-protector-strong} causes GCC to emit code that provides
some protection against some classes of attacks based on buffer overflows.
One could rightly argue code should simply avoid ever overflowing buffers,
but things like memory allocators and userspace memory accesses must
use code that can be difficult to distinguish from buffer overflows.
It is not clear that the ever-increasing variety of attacks should
affect the standard.

The \co{-fno-strict-overflow} causes GCC to act as if signed integer
overflow is defined behavior,
which also affects the definition of ``semantic dependency''.
This might be a controversial choice, and another option would be to add
a new set of signed integer types to the standard for which overflow is
defined as wrapping, similar to the situation with unsigned integers.

The \co{-Werror=strict-prototypes} causes GCC to warn if old-style
non-ANSI function prototypes are used.
This helps avoid certain classes of bugs.
Warnings are by design outside the scope of the standard.

To sum up, user preference can exert a nontrivial influence over
language semantics, and in particular can affect some aspects of the
definition of ``semantic dependency''.

@@

One entirely reasonable reaction is that, given the issues raised in
Sections~\ref{sec:Properties of Semantic Dependencies}
and~\ref{sec:The Fundamental Property of Semantic Dependencies},
load/store reordering is the least of the problems faced by tooling.
Nevertheless, this appendix expands on
Section~\ref{sec:Semantic Dependencies and Tooling}
by looking into the possibility of:
\begin{itemize}
\item	Tooling focusing on only part of the language,
\item	Changing the language to eliminate some aspects that are
	troublesome for tooling (and potentially accepting performance
	and energy-efficiency shortfalls), and
\item	Changing the language to better delineate those portions that
	tooling can easily accommodate.
\end{itemize}

@@

Both the ARM and PowerPC architectural memory models permit reordering prior loads
against subsequent stores, but actual tests on recent hardware fail to
produce any evidence of any such reordering.
Nevertheless, thus far neither ARM nor PowerPC hardware architects
have been willing to strengthen their memory models so as to forbid
such reordering.
% @@@ Cite Grisenthwaite

@@

However, if the load suffers a cache miss, address translation will have
completed long before the load returns its value, meaning that a later
store might well commit before the load completes.
In fact, that later store might commit before the store which supplies the
value returned by the load!
Which explains why some hardware systems really do reorder prior loads
and later stores.

@@

However, some kernels and applications have ways of handling even
uncorrectable errors.
Operating-system kernels encountering an ECC error might note that the
affected data was being used by only one user process, and might react by
killing that user process, taking care to account for memory shared with
other processes.
Some user applications might note that the corrupted data affected only
a particular computation, and might react by restarting that computation.
In both cases, it is only reasonable to react to an uncorrectable ECC
error in a load from a read-only mapped file by re-reading that data
from that file, then restarting that load instruction.
In all cases, the kernel and the user application would need to continue
execution, and would thus require an exact exception.

@@

If you would like an authoritative answer to this question, you should of
course ask your friendly local hardware architect.
In the meantime, here is some semi-informed speculation on this topic:

\begin{itemize}
\item	Some hardware might choose to forego ECC, for example, in order to
	reduce cost for low-end systems or to improve energy efficiency
	for battery-powered systems.
\item	ECC error correction might still be done in hardware, avoiding
	the need for precise software-visible exceptions.
\item	Some systems might prefer to immediately shut down in response to
	an uncorrectable ECC error, perhaps due to safety considerations.
	This would entirely avoid the need for software-visible
	ECC-related exceptions.\footnote{
		One motivation for immediate shutdown is that if there
		are uncorrectable errors, there might soon be errors
		that translate one valid bit pattern to another valid
		bit pattern, which could result in a lack of safety.}
\item	Someone might come up with a clever way of correcting ECC errors
	in firmware, avoiding the need for precise software-visible
	exceptions.
\item	As late as early 2024, some GPGPUs have been observed reordering
	earlier loads against later stores.
	Other hardware architects might therefore feel the need to keep
	this option open for their own systems.
\end{itemize}

@@

However, this would mean that programs compiled in this way would not
be guaranteed to run correctly on weakly ordered systems that lack
ECC (or that lack precise exceptions).
In addition, this change is more strict than necessary, forbidding
optimizations that tooling could in fact analyze.
Therefore, the next section looks at standardizing a less severe
restriction.
